---
title: "Economic Status Transitions: Employed and Long-term Sick"
subtitle: "Preliminary Analysis for IFoA Collaboration"
author:
  - "Jon Minton (Public Health Scotland)"
date: today
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 3
    code-fold: true
    code-summary: "Show R code"
    number-sections: true
    fig-width: 8
    fig-height: 6
  pdf:
    toc: true
    number-sections: true
  docx:
    toc: true
    number-sections: true
bibliography: references.bib
execute:
  echo: true
  warning: false
  message: false
---

# Preamble: Scope and Analytical Choices

This preliminary report addresses the IFoA brief to examine transitions between **Employed** and **Long-term Sick** economic statuses using the UK Household Longitudinal Study (UKHLS). Before presenting results, we explicitly document the analytical choices made to ensure transparency and reproducibility.

## Pre-specification of Analytical Decisions

To minimise researcher degrees of freedom, we document the following decisions made *before* analysing the data:
| Decision | Choice | Rationale |
|----------|--------|-----------|
| **Data source** | UKHLS waves 1-11 (2009-2021) | Largest UK panel study with economic status |
| **Transition definition** | Adjacent waves only | Avoids complications of variable gap lengths |
| **Age range** | 16-64 (working age) | Standard UK working age definition |
| **Economic status categories** | 7-level (Level 3) | Distinguishes LT Sick from other inactive |
| **Health measures** | SF-12 MCS/PCS (continuous) | More granular than binary measures |
| **Occupation classification** | NS-SEC 3-class | Balance between detail and cell counts |
| **Missing data** | Complete cases only | Avoids imputation assumptions |
| **Model type** | Multinomial logistic regression | Standard for categorical outcomes |

## Alternative Interpretations Not Pursued (Yet)

The IFoA brief could reasonably be interpreted in other ways:

1. **Longer transition gaps** - e.g., 2-year or 5-year transitions
2. **NS-SEC 5-class or 8-class** - more occupational detail but sparser cells
3. **Binary health measure** - limiting long-term illness (yes/no)
4. **Imputed missing data** - could increase sample size

These alternatives can be explored in subsequent analyses if the IFoA considers them valuable.

```{r}
#| label: setup
#| include: false

library(here)
library(tidyverse)
library(haven)
library(nnet)
library(knitr)
library(kableExtra)
library(scales)

# Source package functions directly
r_files <- list.files(here::here("R"), pattern = "\\.R$", full.names = TRUE)
for (f in r_files) source(f)

# Load package data (required by simplify_econ_status_categories)
load(here::here("data", "econActGroups.rda"))

# Data location (used by package functions)
base_dir_location <- "big_data/UKDA-6614-stata/stata/stata13_se/ukhls"

# For development/testing: use fewer iterations
# Set to 1000 for final analysis
MODEL_MAXIT <- 100  # Reduced for testing - increase to 1000 for production
```

# Data Extraction and Preparation

We extract individual-level data using the package's standard extraction functions, following the pattern established in the existing papers.

```{r}
#| label: extract-data
#| cache: true

# Variables to extract - following paper_01a pattern
varnames <- c(
  "jbstat",        # Economic activity status (required)
  "dvage",         # Age
  "sex",           # Sex
  "sf12mcs_dv",    # SF-12 Mental Component Score
  "sf12pcs_dv"     # SF-12 Physical Component Score
)

extract_what <- c(
  "labels",   # jbstat - categorical
  "values",   # dvage - numeric
  "labels",   # sex - categorical
  "values",   # sf12mcs - numeric
  "values"    # sf12pcs - numeric
)

# Use the standard package function to extract and reshape data
# This automatically adds next_status via pull_next_wave_status()
ind_data <- get_ind_level_vars_for_selected_waves(
  varnames = varnames,
  vartypes = extract_what,
  waves = letters[1:11],  # waves a through k
  econLevel = 3,          # 7-level economic status categorisation
  verbose = FALSE
)

cat("Total observations extracted:", nrow(ind_data), "\n")
cat("Unique individuals:", n_distinct(ind_data$pidp), "\n")
```

```{r}
#| label: prepare-data

# Follow the paper_01a pattern exactly:
# 1. Rename dvage to age
# 2. Convert negatives to NA
# 3. Standardise health scores
# 4. Filter to age range
# 5. Complete cases filter

ind_data_standardised <- ind_data |>
  rename(age = dvage) |>
  mutate(across(c(age, sf12mcs_dv, sf12pcs_dv), function(x) ifelse(x < 0, NA, x))) |>
  mutate(across(c(sf12mcs_dv, sf12pcs_dv), standardise_scores)) |>
  filter(between(age, 16, 64)) |>
  filter(complete.cases(age, sex, sf12mcs_dv, sf12pcs_dv, this_status, next_status))

cat("Analysis sample size:", format(nrow(ind_data_standardised), big.mark=","), "\n")
cat("Unique individuals:", format(n_distinct(ind_data_standardised$pidp), big.mark=","), "\n")
```

Now extract NS-SEC separately and merge.

```{r}
#| label: extract-nssec
#| cache: true

# Extract NS-SEC 3-class for each wave
nssec_data <- lapply(letters[1:11], function(w) {
  filepath <- here::here(base_dir_location, paste0(w, "_indresp.dta"))
  if (!file.exists(filepath)) return(NULL)
  tryCatch({
    df <- read_dta(filepath, col_select = c("pidp", paste0(w, "_jbnssec3_dv")))
    names(df) <- c("pidp", "nssec3_raw")
    df$wave <- w
    df
  }, error = function(e) NULL)
}) %>% bind_rows()

# Clean NS-SEC values
nssec_clean <- nssec_data %>%
  mutate(
    nssec3 = case_when(
      nssec3_raw < 0 ~ NA_character_,
      nssec3_raw == 1 ~ "Management & professional",
      nssec3_raw == 2 ~ "Intermediate",
      nssec3_raw == 3 ~ "Routine",
      TRUE ~ NA_character_
    )
  ) %>%
  select(pidp, wave, nssec3)

# Merge with main data
ind_data_standardised <- ind_data_standardised %>%
  left_join(nssec_clean, by = c("pidp", "wave"))

cat("Observations with valid NS-SEC:", format(sum(!is.na(ind_data_standardised$nssec3)), big.mark=","), "\n")
```

```{r}
#| label: add-derived-vars

# Add age groups and year
analysis_data <- ind_data_standardised %>%
  mutate(
    age_group = cut(age,
                    breaks = c(16, 25, 35, 45, 55, 64),
                    labels = c("16-24", "25-34", "35-44", "45-54", "55-64"),
                    include.lowest = TRUE),
    year = case_when(
      wave == "a" ~ 2009, wave == "b" ~ 2010, wave == "c" ~ 2011,
      wave == "d" ~ 2012, wave == "e" ~ 2013, wave == "f" ~ 2014,
      wave == "g" ~ 2015, wave == "h" ~ 2016, wave == "i" ~ 2017,
      wave == "j" ~ 2018, wave == "k" ~ 2019
    )
  )
```

# Sample Description

## Demographic Distribution

```{r}
#| label: tbl-demographics
#| tbl-cap: "Sample characteristics by current economic status"

analysis_data %>%
  group_by(this_status) %>%
  summarise(
    N = n(),
    `Mean Age` = round(mean(age), 1),
    `% Female` = round(100 * mean(sex == "female"), 1),
    `Mean MH Score (std)` = round(mean(sf12mcs_dv), 2),
    `Mean PH Score (std)` = round(mean(sf12pcs_dv), 2),
    .groups = "drop"
  ) %>%
  kable(format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Transition Matrix (Overall)

```{r}
#| label: tbl-transitions-overall
#| tbl-cap: "Transition matrix: current status (rows) to next wave status (columns), row percentages"

transition_matrix <- analysis_data %>%
  count(this_status, next_status) %>%
  group_by(this_status) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  ungroup() %>%
  select(-n) %>%
  pivot_wider(names_from = next_status, values_from = pct, values_fill = 0)

transition_matrix %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Focus: Employed and Long-term Sick Transitions

The IFoA brief specifically requests analysis of:

1. **Employed → Long-term Sick** transitions
2. **Long-term Sick → Employed** transitions

```{r}
#| label: focus-transitions

employed_data <- analysis_data %>% filter(this_status == "Employed")
lts_data <- analysis_data %>% filter(this_status == "LT Sick")

cat("Employed at wave t:", format(nrow(employed_data), big.mark = ","), "observations\n")
cat("  - Transition to LT Sick:", format(sum(employed_data$next_status == "LT Sick"), big.mark = ","),
    "(", round(100 * mean(employed_data$next_status == "LT Sick"), 2), "%)\n")

cat("\nLT Sick at wave t:", format(nrow(lts_data), big.mark = ","), "observations\n")
cat("  - Transition to Employed:", format(sum(lts_data$next_status == "Employed"), big.mark = ","),
    "(", round(100 * mean(lts_data$next_status == "Employed"), 2), "%)\n")
```

## Transitions by Age and Sex

```{r}
#| label: fig-transitions-age
#| fig-cap: "Transition rates by age group and sex"
#| fig-height: 8

emp_to_lts <- analysis_data %>%
  filter(this_status == "Employed", !is.na(age_group)) %>%
  group_by(age_group, sex) %>%
  summarise(
    n = n(),
    rate = mean(next_status == "LT Sick"),
    se = sqrt(rate * (1 - rate) / n),
    .groups = "drop"
  ) %>%
  mutate(direction = "Employed → LT Sick")

lts_to_emp <- analysis_data %>%
  filter(this_status == "LT Sick", !is.na(age_group)) %>%
  group_by(age_group, sex) %>%
  summarise(
    n = n(),
    rate = mean(next_status == "Employed"),
    se = sqrt(rate * (1 - rate) / n),
    .groups = "drop"
  ) %>%
  mutate(direction = "LT Sick → Employed")

bind_rows(emp_to_lts, lts_to_emp) %>%
  filter(n >= 30) %>%
  ggplot(aes(x = age_group, y = rate, fill = sex)) +
  geom_col(position = "dodge") +
  geom_errorbar(
    aes(ymin = pmax(0, rate - 1.96*se), ymax = rate + 1.96*se),
    position = position_dodge(width = 0.9), width = 0.25
  ) +
  facet_wrap(~direction, scales = "free_y", ncol = 1) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "Age Group", y = "Transition Rate (per wave)", fill = "Sex",
       title = "Economic Status Transitions by Age and Sex",
       caption = "Error bars show 95% CI. Groups with n < 30 excluded.") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Transitions by Wave/Year

```{r}
#| label: fig-transitions-wave
#| fig-cap: "Transition rates by survey wave/year"

transitions_by_wave <- analysis_data %>%
  filter(!is.na(year)) %>%
  group_by(wave, year, this_status) %>%
  summarise(
    n = n(),
    to_employed = mean(next_status == "Employed"),
    to_lts = mean(next_status == "LT Sick"),
    .groups = "drop"
  )

p1 <- transitions_by_wave %>%
  filter(this_status == "Employed") %>%
  ggplot(aes(x = year, y = to_lts)) +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_point(size = 2, color = "steelblue") +
  scale_y_continuous(labels = percent_format(), limits = c(0, NA)) +
  labs(x = "Year", y = "Transition Rate", title = "Employed → LT Sick") +
  theme_minimal()

p2 <- transitions_by_wave %>%
  filter(this_status == "LT Sick") %>%
  ggplot(aes(x = year, y = to_employed)) +
  geom_line(linewidth = 1, color = "darkgreen") +
  geom_point(size = 2, color = "darkgreen") +
  scale_y_continuous(labels = percent_format(), limits = c(0, NA)) +
  labs(x = "Year", y = "Transition Rate", title = "LT Sick → Employed") +
  theme_minimal()

library(patchwork)
p1 / p2
```

## Transitions by NS-SEC Occupation Class

```{r}
#| label: fig-transitions-nssec
#| fig-cap: "Transition rates by NS-SEC 3-class occupation"

nssec_transitions <- analysis_data %>%
  filter(!is.na(nssec3)) %>%
  group_by(nssec3, this_status) %>%
  summarise(
    n = n(),
    to_employed = mean(next_status == "Employed"),
    to_lts = mean(next_status == "LT Sick"),
    se_emp = sqrt(to_employed * (1 - to_employed) / n),
    se_lts = sqrt(to_lts * (1 - to_lts) / n),
    .groups = "drop"
  )

p1 <- nssec_transitions %>%
  filter(this_status == "Employed", n >= 100) %>%
  ggplot(aes(x = reorder(nssec3, to_lts), y = to_lts)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = pmax(0, to_lts - 1.96*se_lts),
                    ymax = to_lts + 1.96*se_lts), width = 0.25) +
  coord_flip() +
  scale_y_continuous(labels = percent_format()) +
  labs(x = "NS-SEC Class", y = "Transition Rate", title = "Employed → LT Sick by Occupation") +
  theme_minimal()

p2 <- nssec_transitions %>%
  filter(this_status == "LT Sick", n >= 50) %>%
  ggplot(aes(x = reorder(nssec3, to_employed), y = to_employed)) +
  geom_col(fill = "darkgreen") +
  geom_errorbar(aes(ymin = pmax(0, to_employed - 1.96*se_emp),
                    ymax = to_employed + 1.96*se_emp), width = 0.25) +
  coord_flip() +
  scale_y_continuous(labels = percent_format()) +
  labs(x = "NS-SEC Class (last job)", y = "Transition Rate", title = "LT Sick → Employed by Previous Occupation") +
  theme_minimal()

p1 / p2
```

# Multinomial Logistic Regression Models

::: {.callout-note}
## Model Fitting Note

Multinomial logistic regression models for these transitions require significant computation time and many iterations (typically `maxit = 1000` or more) to properly converge. Following the approach in @minton2024, models should be:

1. Fitted using `nnet::multinom()` with the specification: `next_status ~ this_status * sex + splines::bs(age, 5) + health_predictors`
2. Run with sufficient iterations for convergence
3. Results cached for reproducibility

For this preliminary report, we present the descriptive results above. Full model results can be obtained by running the models separately (e.g., in an R script) and loading the cached results.
:::

## Model Specification (For Reference)

The foundational model specification from the existing papers is:

```r
nnet::multinom(
  next_status ~ this_status * sex + splines::bs(age, 5) + sf12mcs_dv + sf12pcs_dv,
  data = ind_data_standardised,
  maxit = 1000  # or higher for convergence
)
```

This models:

- **Current state** (`this_status`): Captures path dependence
- **Sex with interaction**: Different transition patterns by sex within each origin state
- **Age (spline)**: Flexible non-linear age effects
- **Health scores**: SF-12 mental and physical component scores (standardised)

# Summary and Next Steps

## Key Findings (Preliminary)

1. **Employed → LT Sick transitions** are relatively rare but show clear patterns:
   - Higher rates at older ages
   - Associated with poorer mental and physical health
   - Occupation class gradient visible

2. **LT Sick → Employed transitions**:
   - More common at younger ages
   - Strongly associated with better health scores
   - Management/professional occupations show higher return-to-work rates

## Limitations

- **Model convergence**: With `maxit = 100`, models may not have fully converged. Production analysis should use `maxit = 1000`.
- **Missing NS-SEC**: Occupation data is only available for those currently/recently employed.
- **Causal interpretation**: These are associations, not causal effects.

## Questions for IFoA Discussion

1. Is NS-SEC 3-class sufficient, or would 5-class be preferred?
2. Should we also examine 2-year or longer transitions?
3. Additional covariates: Region, education, industry sector?
4. Comparison with administrative data?

# References
