---
title: "Economic Status Transitions: Employed ↔ Long-term Sick"
subtitle: "IFoA Collaboration: Model Results and Counterfactual Analysis"
author:
  - "Jon Minton"
date: today
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 3
    code-fold: true
    code-summary: "Show R code"
    number-sections: true
    fig-width: 8
    fig-height: 6
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
  docx:
    toc: true
    number-sections: true
bibliography: references.bib
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(here)
library(tidyverse)
library(nnet)
library(knitr)
library(kableExtra)
library(scales)
library(patchwork)
library(splines)

# Source package functions
r_files <- list.files(here::here("R"), pattern = "\\.R$", full.names = TRUE)
for (f in r_files) source(f)
load(here::here("data", "econActGroups.rda"))

# Pre-compute PAF for executive summary
# (Full analysis details shown later in document)
best_model <- readRDS(here::here("reports/ifoa/models/mod_06_health_nssec_year.rds"))
data_full <- readRDS(here::here("reports/ifoa/models/data_full.rds"))

# Define ordinal NS-SEC levels (routine/manual → intermediate → management/professional)
nssec_levels <- c("Routine & Manual", "Intermediate", "Management & Professional")

# Get NS-SEC distribution from employed 45-55 year olds
nssec_dist <- data_full |>
  filter(this_status == "Employed", between(age, 45, 55)) |>
  mutate(nssec3 = factor(nssec3, levels = nssec_levels)) |>
  count(sex, nssec3) |>
  group_by(sex) |>
  mutate(prop = n / sum(n)) |>
  ungroup()

# Simulation function
simulate_transitions <- function(model, n_per_sex = 10000, age = 50,
                                  nssec_distribution, wave = 11,
                                  counterfactual_nssec = NULL) {
  results <- list()
  for (s in c("male", "female")) {
    sex_dist <- nssec_distribution |> filter(sex == s)
    if (is.null(counterfactual_nssec)) {
      n_by_nssec <- round(sex_dist$prop * n_per_sex)
      n_by_nssec[1] <- n_by_nssec[1] + (n_per_sex - sum(n_by_nssec))
      cohort <- tibble(sex = s, nssec3 = rep(sex_dist$nssec3, times = n_by_nssec))
    } else {
      cohort <- tibble(sex = s, nssec3 = rep(counterfactual_nssec, n_per_sex))
    }
    cohort <- cohort |>
      mutate(this_status = "Employed", age = age, sf12mcs_std = 0, sf12pcs_std = 0, wave_num = wave)
    preds <- predict(model, newdata = cohort, type = "probs")
    cohort$prob_ltsick <- preds[, "Inactive long term sick"]
    results[[s]] <- cohort |>
      summarise(sex = first(sex), n = n(), expected_ltsick = sum(prob_ltsick), mean_prob = mean(prob_ltsick))
  }
  bind_rows(results)
}

baseline <- simulate_transitions(best_model, 10000, 50, nssec_dist, 11)
counterfactual <- simulate_transitions(best_model, 10000, 50, nssec_dist, 11, "Management & Professional")

paf_results <- baseline |>
  rename(baseline_expected = expected_ltsick, baseline_prob = mean_prob) |>
  left_join(
    counterfactual |>
      rename(counter_expected = expected_ltsick, counter_prob = mean_prob) |>
      select(sex, counter_expected, counter_prob),
    by = "sex"
  ) |>
  mutate(
    diff_absolute = baseline_expected - counter_expected,
    diff_relative_pct = (baseline_expected - counter_expected) / baseline_expected * 100,
    paf = diff_relative_pct
  )

paf_overall <- round(mean(paf_results$paf), 0)
```

# Executive Summary

This report presents multinomial logistic regression models for economic status transitions, focusing on movements between **Employed** and **Long-term Sick** states. Rising economic inactivity due to long-term sickness has become a significant policy concern in the UK [@tinson2022]. Using data from the UK Household Longitudinal Study (UKHLS) waves 1-11 (2009-2020) [@platt2021], we estimate the role of health (SF-12 mental and physical component scores [@ware1996]) and occupation class (NS-SEC 3-class [@nssec]) in explaining these transitions.

**Key findings:**

1. **Model selection**: The best-fitting model (by BIC) includes health scores, occupation class, and a year trend, without interaction terms
2. **Health effects**: Both mental and physical health strongly predict transitions - poorer health increases risk of becoming long-term sick
3. **Occupation gradient**: Routine/manual occupations have higher Employed → LT Sick transition rates than management/professional occupations
4. **Population Attributable Fraction**: Approximately `r paf_overall`% of Employed → LT Sick transitions among 50-year-olds can be attributed to occupational class inequalities

# Analytical Framework

{{< include _analytical_decisions.qmd >}}

# Data Description

```{r}
#| label: data-description

# Sample sizes by dataset
data_core <- readRDS(here::here("reports/ifoa/models/data_core.rds"))
data_health <- readRDS(here::here("reports/ifoa/models/data_health.rds"))

tibble(
  Dataset = c("Core (demography only)", "With health scores", "With health + occupation"),
  `N observations` = c(nrow(data_core), nrow(data_health), nrow(data_full)),
  `N individuals` = c(n_distinct(data_core$pidp), n_distinct(data_health$pidp), n_distinct(data_full$pidp))
) |>
  mutate(across(where(is.numeric), ~comma(.))) |>
  kable(caption = "Sample sizes by model dataset") |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Economic Status Distribution

The seven-level economic status categorisation distinguishes between types of economic inactivity:

```{r}
#| label: status-distribution

data_full |>
  count(this_status, name = "N") |>
  mutate(Percent = percent(N / sum(N), accuracy = 0.1)) |>
  arrange(desc(N)) |>
  kable(col.names = c("Current Economic Status", "N", "Percent"),
        caption = "Distribution of current economic status in analysis sample") |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Transition Matrix

```{r}
#| label: transition-matrix

data_full |>
  count(this_status, next_status) |>
  group_by(this_status) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  ungroup() |>
  select(-n) |>
  pivot_wider(names_from = next_status, values_from = pct, values_fill = 0) |>
  kable(caption = "Transition matrix: row percentages (current → next wave status)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 11) |>
  scroll_box(width = "100%")
```

# Model Comparison

Eight candidate model specifications were fitted and compared using AIC and BIC criteria. Models were fit separately in a script (`_fit_models.R`) and saved for reproducibility.

```{r}
#| label: load-models

# Load model comparison results (best_model and data_full already loaded in setup)
model_comparison <- readRDS(here::here("reports/ifoa/models/model_comparison.rds"))
```

## Model Specifications Tested

| Model | Formula | N | Description |
|-------|---------|---|-------------|
| 1 | `next_status ~ this_status * sex + bs(age, 5)` | 305,327 | Foundation (demography only) |
| 2 | + `wave_num` | 305,327 | + Year trend |
| 3 | + `sf12mcs_std + sf12pcs_std` | 269,786 | + Health scores |
| 4 | + health + `wave_num` | 269,786 | + Health + year |
| 5 | + health + `nssec3` | 190,465 | + Health + occupation |
| 6 | + health + occupation + `wave_num` | 190,465 | + Health + occupation + year |
| 7 | + health × occupation | 190,465 | + Interaction terms |
| 8 | Full (all interactions + year) | 190,465 | Most complex |

## AIC/BIC Comparison

```{r}
#| label: model-comparison-table

model_comparison |>
  select(name, n, converged, aic, bic) |>
  mutate(
    name = str_remove(name, "mod_0") |> str_replace("_", ": "),
    aic = comma(round(aic, 0)),
    bic = comma(round(bic, 0)),
    n = comma(n)
  ) |>
  arrange(bic) |>
  kable(
    col.names = c("Model", "N", "Converged", "AIC", "BIC"),
    caption = "Model comparison ranked by BIC (lower is better)"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover")) |>
  row_spec(1, bold = TRUE, background = "#e6f3ff")
```

**Model selection decision**: Model 6 (`mod_06_health_nssec_year`) is selected as it has the lowest BIC. This model includes:

- Current economic status (`this_status`) with sex interaction
- Age (B-spline with 5 degrees of freedom)
- Standardised SF-12 mental and physical component scores
- NS-SEC 3-class occupation
- Year trend

The interaction models (7 and 8) did not improve BIC despite their additional complexity, indicating that the effect of health on transitions does not differ meaningfully by occupation class.

# Transition Probabilities by NS-SEC

Using the selected model, we estimate transition probabilities for individuals with specific characteristics.

## Employed → Long-term Sick Transitions

We focus on the transition of interest: movement from Employed to Long-term Sick status. We examine how this varies by occupation class, holding other factors constant.

```{r}
#| label: emp-to-ltsick-by-nssec

# Create prediction grid for Employed individuals aged 50
pred_grid_emp <- expand_grid(
  this_status = "Employed",
  sex = c("male", "female"),
  age = 50,
  sf12mcs_std = 0,  # Average mental health
  sf12pcs_std = 0,  # Average physical health
  nssec3 = nssec_levels,  # Use ordinal ordering
  wave_num = 11     # Latest wave (2019-2020)
)

# Get predictions
pred_emp <- predict(best_model, newdata = pred_grid_emp, type = "probs")

# Combine with grid and apply factor ordering
pred_emp_df <- bind_cols(pred_grid_emp, as_tibble(pred_emp)) |>
  select(sex, nssec3, `Inactive long term sick`) |>
  rename(prob_ltsick = `Inactive long term sick`) |>
  mutate(nssec3 = factor(nssec3, levels = nssec_levels))

# Display table
pred_emp_df |>
  mutate(
    prob_ltsick_pct = percent(prob_ltsick, accuracy = 0.01),
    prob_ltsick_per_10k = comma(prob_ltsick * 10000, accuracy = 1)
  ) |>
  select(sex, nssec3, prob_ltsick_pct, prob_ltsick_per_10k) |>
  kable(
    col.names = c("Sex", "NS-SEC Class", "Probability", "Per 10,000"),
    caption = "Predicted Employed → LT Sick transition probability (age 50, average health, wave 11)"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r}
#| label: plot-emp-to-ltsick
#| fig-cap: "Predicted probability of Employed → LT Sick transition by occupation class (age 50, average health)"

pred_emp_df |>
  ggplot(aes(x = nssec3, y = prob_ltsick, fill = sex)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  scale_fill_brewer(palette = "Set1", name = "Sex") +
  labs(
    x = "NS-SEC Occupation Class",
    y = "Probability of Transition to LT Sick",
    title = "Employed → Long-term Sick Transitions",
    subtitle = "Predicted for age 50, average health scores, wave 11 (2019-20)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

# Counterfactual Analysis: Population Attributable Fraction

## Analytical Approach

To quantify the contribution of occupational class to Employed → LT Sick transitions, we conduct a counterfactual simulation:

1. **Baseline scenario**: 10,000 males aged 50 and 10,000 females aged 50, all currently Employed, distributed across NS-SEC classes proportionally to the observed sample
2. **Counterfactual scenario**: Same population, but everyone assigned to "Management & Professional" (the class with lowest LT Sick transition risk)
3. **PAF calculation**: (Baseline transitions - Counterfactual transitions) / Baseline transitions

```{r}
#| label: nssec-distribution

# Display NS-SEC distribution (computed in setup)
nssec_dist |>
  mutate(prop_pct = percent(prop, accuracy = 0.1)) |>
  select(sex, nssec3, n, prop_pct) |>
  kable(
    col.names = c("Sex", "NS-SEC Class", "N", "Proportion"),
    caption = "NS-SEC distribution among employed individuals aged 45-55"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Results: Occupation-Attributable Transitions

```{r}
#| label: paf-table

paf_results |>
  mutate(
    baseline_expected = round(baseline_expected, 1),
    counter_expected = round(counter_expected, 1),
    diff_absolute = round(diff_absolute, 1),
    paf = paste0(round(paf, 1), "%")
  ) |>
  select(sex, n, baseline_expected, counter_expected, diff_absolute, paf) |>
  kable(
    col.names = c("Sex", "N", "Baseline Expected", "Counterfactual Expected",
                  "Difference", "PAF"),
    caption = "Population Attributable Fraction for Employed → LT Sick transitions due to occupational class inequality"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r}
#| label: plot-paf
#| fig-cap: "Expected Employed → LT Sick transitions under baseline and counterfactual scenarios"

paf_plot_data <- paf_results |>
  select(sex, baseline_expected, counter_expected) |>
  pivot_longer(
    cols = c(baseline_expected, counter_expected),
    names_to = "scenario",
    values_to = "expected_transitions"
  ) |>
  mutate(
    scenario = case_when(
      scenario == "baseline_expected" ~ "Baseline\n(observed NS-SEC distribution)",
      scenario == "counter_expected" ~ "Counterfactual\n(all Management & Professional)"
    )
  )

ggplot(paf_plot_data, aes(x = scenario, y = expected_transitions, fill = sex)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = round(expected_transitions, 0)),
            position = position_dodge(width = 0.9), vjust = -0.5) +
  scale_fill_brewer(palette = "Set1", name = "Sex") +
  labs(
    x = "",
    y = "Expected Transitions to LT Sick (per 10,000)",
    title = "Effect of Equalising Occupational Class",
    subtitle = "Among 10,000 employed individuals aged 50"
  ) +
  theme_minimal()
```

## Interpretation

Among a hypothetical cohort of 10,000 employed 50-year-olds with average health:

- **Baseline**: With the observed distribution of occupational classes, we expect approximately `r round(paf_results$baseline_expected[paf_results$sex == "male"], 0)` males and `r round(paf_results$baseline_expected[paf_results$sex == "female"], 0)` females to transition to long-term sickness over the following year.

- **Counterfactual**: If everyone had the transition risk of Management & Professional occupations, we would expect approximately `r round(paf_results$counter_expected[paf_results$sex == "male"], 0)` males and `r round(paf_results$counter_expected[paf_results$sex == "female"], 0)` females to make this transition.

- **PAF**: Approximately `r round(mean(paf_results$paf), 0)`% of Employed → LT Sick transitions among 50-year-olds can be attributed to occupational class inequality.

This suggests that if all workers had the same low transition risk as those in management and professional occupations, we could expect roughly `r round(mean(paf_results$paf), 0)`% fewer transitions from employment to long-term sickness in this age group.

# Time Trends

The model includes a year trend (`wave_num`). This captures systematic changes in transition probabilities over the 2009-2020 period.

```{r}
#| label: time-trend
#| fig-cap: "Predicted Employed → LT Sick probability over time (age 50, average health, Routine & Manual)"

# Predictions across waves
time_grid <- expand_grid(
  this_status = "Employed",
  sex = c("male", "female"),
  age = 50,
  sf12mcs_std = 0,
  sf12pcs_std = 0,
  nssec3 = "Routine & Manual",
  wave_num = 1:11
)

time_preds <- predict(best_model, newdata = time_grid, type = "probs")
time_df <- bind_cols(time_grid, as_tibble(time_preds)) |>
  mutate(
    year = 2008 + wave_num,
    prob_ltsick = `Inactive long term sick`
  )

ggplot(time_df, aes(x = year, y = prob_ltsick, colour = sex)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  scale_colour_brewer(palette = "Set1", name = "Sex") +
  scale_x_continuous(breaks = seq(2009, 2019, 2)) +
  labs(
    x = "Year",
    y = "Probability of Transition to LT Sick",
    title = "Time Trend in Employed → LT Sick Transitions",
    subtitle = "Routine & Manual occupations, age 50, average health"
  ) +
  theme_minimal()
```

# Discussion

## Key Findings

1. **Health is the strongest predictor**: Both mental and physical health scores substantially improve model fit (AIC drops from ~289k to ~244k when health is added).

2. **Occupation class matters**: Adding NS-SEC occupation further improves fit. Those in routine/manual occupations have higher risk of transitioning from employment to long-term sickness.

3. **No evidence for interaction effects**: The relationship between health and transitions does not differ significantly by occupation class (interaction models did not improve BIC).

4. **Year trend is significant**: Transition probabilities have changed systematically over 2009-2020, though the magnitude is modest.

5. **PAF for occupation**: Approximately `r round(mean(paf_results$paf), 0)`% of Employed → LT Sick transitions among 50-year-olds can be attributed to occupational class inequalities.

## Limitations

- **Pre-pandemic data**: Analysis uses waves 1-11 (2009-2020). Post-pandemic patterns may differ.
- **Complete cases**: Individuals missing health or occupation data are excluded, reducing sample size.
- **Causal interpretation**: These are associations controlling for observed confounders; unobserved confounding may remain.
- **NS-SEC measurement**: Based on current/recent job; may not capture full occupational history.

## Implications for Actuarial Analysis

1. **Risk stratification**: Occupation class provides meaningful stratification of transition risk beyond health status alone.

2. **Projection scenarios**: The model can be used to project future LT Sick populations under different assumptions about health trajectories or occupational composition.

3. **Intervention targeting**: The PAF analysis suggests that reducing occupational class gradients in transition risk could meaningfully reduce the flow into long-term sickness.

# Technical Appendix

## Model Coefficients

```{r}
#| label: model-coefficients

# Extract and display coefficients for key transitions
coefs <- coef(best_model)

# Focus on Employed → LT Sick (relative to staying Employed)
if ("Inactive long term sick" %in% rownames(coefs)) {
  ltsick_coefs <- coefs["Inactive long term sick", ]

  # Select key coefficients
  key_coefs <- ltsick_coefs[c(
    "(Intercept)",
    "this_statusEmployed:sexmale",
    "sf12mcs_std",
    "sf12pcs_std",
    "nssec3Intermediate",
    "nssec3Routine & Manual",
    "wave_num"
  )]

  tibble(
    Coefficient = names(key_coefs),
    Estimate = round(key_coefs, 3),
    `Odds Ratio` = round(exp(key_coefs), 3)
  ) |>
    kable(caption = "Selected coefficients for Employed → LT Sick transition (reference: Employed, female, Management & Professional)") |>
    kable_styling(bootstrap_options = c("striped", "hover"))
}
```

## Session Information

```{r}
#| label: session-info
sessionInfo()
```

# References
